\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-07-27 papers/methods/paper.tex
Adding a bit more clarification from dans emails

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed Type Ia supernovae (SNIa) have increased by more than an order of magnitude, with contributions from modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, Sako2014}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structureand the CMB. For a recent review, see \citet{Huterer2018}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

In attempts to tease out the nature of dark energy, active and planned surveys are once again ramping up their statistical power. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has observed thousands of Type Ia supernova, attaining both spectroscopic and photometric selected samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will observe scores of thousands of photometrically classified supernovae. Such increased statistical power demands greater fidelity and flexibility in modelling the supernovae for cosmological purposes, as systematic uncertainties will prove to be the limiting factor in our analyses \citep{Betoule2014}.

@@

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about correlation between sources of uncertainty. For example, the underlying supernova colour population's mean and skewness is highly correlated, however this correlation is lost when determining population uncertainty using numerical derivatives of population permutations. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. A common solution to this computational cost problem is to bin the supernova.

@@

Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift, and subtract this shift from our observed data. With the extra dispersion introduced from the $\kappa$ parameters addressing the under-reporting of error, shifting the mean colour address the non-Gaussianity introduced by chromatic smearing. We allow the amount of shift to be parametrised to account for different intrinsic scatter models where parameter $s_c$ controls the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations. The scatter models, sourced from \citet{Guy2010} and \citet{Chotard2011} are discussed in detail in Section \ref{sec:simdes}. The maximum shift was found to using the intrinsic scatter model from \citet{Chotard2011}. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a global shift in supernovae colour across all surveys. We place a uniform prior on $s_c$. It is important to note that the effects being modelled and characterised by simulations here are features of the supernova population, not cosmological features, and so the choice of fiducial cosmology has negligible impact.

@@

 In addition to the improvements in the scatter models over the simple data, we also include peculiar velocities for the LowZ sample, and our full treatment of systematics. Our simulated populations are sourced from \citet[][hereafter {\sk}]{Scolnic2016}. The selection effects were quantified by comparing all the generated supernovae to those that pass our simulated cuts, as shown in Figure~\ref{fig:sf_snana}. It is from this simulation that our analytic determination of the selection functions for the LowZ and DES survey are based. We note that there was no significant difference in the selection effect of Malmquist bias between using the {\gten} or {\celeven} scatter model compared to the fit uncertainty.

@@

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. The quoted value for $w$ represents the average mean of the fits, with the uncertainty being the average standard deviation of the fits. In brackets, the simulation scatter (the standard deviation of the mean of 100 fits) is shown, and the bias significance represents our confidence that the deviation in the mean $w$ away from $-1$ is not due to statistical fluctuation. With systematics enabled, both the {\gten} and {\celeven} models show evidence of bias, scattering to either side of the simulated value of $w=-1$. However, their deviation from the truth value represents a shift of approximately $0.3\sigma$ when taking into account the uncertainty on fits to $w$. {\green The bias is sub-dominant to both the size of the uncertainty for each fit, and the scatter induced by statistical variance in the simulations. We also note that the simulations do not vary cosmological parameters nor population, and as our model does include uncertainty on those values, the simulation scatter is expected to be less than the model uncertainty, and in fact represents a minimum bound on permissible uncertainty values.}}
	\label{tab:bulk_summary}
	\begin{tabular}{l|cc}
		\hline
		Model & $w$ (scatter) & Bias Significance \\
		\hline
		{\gten} Stat + Syst  &  $-1.03\pm 0.10\ (0.08)$ & $3.9\sigma$  \\
		{\celeven} Stat + Syst  &  $-0.97\pm 0.09\ (0.06)$ & $4.4\sigma$  \\
		{\gten}  Stat         &  $-1.01\pm 0.08\ (0.08)$ & $2.0\sigma$   \\
		{\celeven} Stat         &  $-0.96\pm 0.07\ (0.05)$ & $7.9\sigma$   \\
		\hline
	\end{tabular}
\end{table}

>>>>

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed Type Ia supernovae (SNIa) have increased by more than an order of magnitude, with contributions from modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, Sako2014}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014, Scolnic2017} have been combined with complimentary probes of large scale structureand the CMB. For a recent review, see \citet{Huterer2018}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

In attempts to tease out the nature of dark energy, active and planned surveys are once again ramping up their statistical power. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has observed thousands of Type Ia supernova, attaining both spectroscopic and photometric selected samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will observe scores of thousands of photometrically classified supernovae. Such increased statistical power demands greater fidelity and flexibility in modelling the supernovae for cosmological purposes, as systematic uncertainties will prove to be the limiting factor in our analyses \citep{Betoule2014, Scolnic2017}.

@@

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about correlation between sources of uncertainty. For example, the underlying supernova colour population's mean and skewness is highly correlated, however this correlation is lost when determining population uncertainty using numerical derivatives of population permutations. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. {\green A common solution to this computational cost problem is to bin the supernova, which then produces a matrix of arbitrary size. Whilst binning data does result in some loss of information, recent works tested against simulations show that this information is not cosmologically significant \citep{Scolnic2016, Kessler2017}.}

@@

Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift. {\green This redshift-dependent array is added to our data-set and is included in our modelling by allowing a controlled subtract this shift from our observed data, where the amount of shift subtracted is dictated by parameter $s_c$. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a global shift in supernovae colour across all surveys. We place a uniform prior on $s_c$.} With the extra dispersion introduced from the $\kappa$ parameters addressing the under-reporting of error, shifting the mean colour address the non-Gaussianity introduced by chromatic smearing. We allow the amount of shift to be parametrised to account for different intrinsic scatter models where parameter $s_c$ controls the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations. The scatter models, sourced from \citet{Guy2010} and \citet{Chotard2011} are discussed in detail in Section \ref{sec:simdes}. The maximum shift was found to using the intrinsic scatter model from \citet{Chotard2011}. It is important to note that the effects being modelled and characterised by simulations here are features of the supernova population, not cosmological features, and so the choice of fiducial cosmology has negligible impact.

@@

 In addition to the improvements in the scatter models over the simple data, we also include peculiar velocities for the LowZ sample, and our full treatment of systematics. {\green Our simulated populations are sourced from \citet[][hereafter {\sk}]{Scolnic2016} and shown in Table \ref{tab:dist}. Initial tests were also done with a second, Gaussian population, however cosmological parameters were not significantly impacted by choice of the underlying population and we continue using only the {\sk} population for computational efficiency.} The selection effects were quantified by comparing all the generated supernovae to those that pass our simulated cuts, as shown in Figure~\ref{fig:sf_snana}. It is from this simulation that our analytic determination of the selection functions for the LowZ and DES survey are based. We note that there was no significant difference in the selection effect of Malmquist bias between using the {\gten} or {\celeven} scatter model compared to the fit uncertainty.

@@

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. The quoted value for $w$ represents the average mean of the fits, with the uncertainty being the average standard deviation of the fits. In brackets, the simulation scatter (the standard deviation of the mean of 100 fits) is shown, and the bias significance represents our confidence that the deviation in the mean $w$ away from $-1$ is not due to statistical fluctuation. With systematics enabled, both the {\gten} and {\celeven} models show evidence of bias, scattering to either side of the simulated value of $w=-1$. However, their deviation from the truth value represents a shift of approximately $0.3\sigma$ when taking into account the uncertainty on fits to $w$. {\green The bias is sub-dominant to both the size of the uncertainty for each fit, and the scatter induced by statistical variance in the simulations. We also note that the simulations do not vary cosmological parameters nor population. As our model does include uncertainty on those values, the simulation scatter is expected to be less than the model uncertainty, and represents a minimum bound on permissible uncertainty values.}}
	\label{tab:bulk_summary}
	\begin{tabular}{l|cc}
		\hline
		Model & $w$ (scatter) & Bias Significance \\
		\hline
		{\gten} Stat + Syst  &  $-1.03\pm 0.10\ (0.08)$ & $3.9\sigma$  \\
		{\celeven} Stat + Syst  &  $-0.97\pm 0.09\ (0.06)$ & $4.4\sigma$  \\
		{\gten}  Stat         &  $-1.01\pm 0.08\ (0.08)$ & $2.0\sigma$   \\
		{\celeven} Stat         &  $-0.96\pm 0.07\ (0.05)$ & $7.9\sigma$   \\
		\hline
	\end{tabular}
\end{table}
